---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mpw

<!-- badges: start -->
<!-- badges: end -->

The goal of mpw is to estimate Marginalized Piecewise Weibull Frailty 
Models in R.

## Installation

You can install the development version of mpw from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("swihart/mpw")
```

## Examples

These are basic examples.  

## PS($\alpha,\alpha,0$) Positive Stable Example

This is the $F_{\alpha}$ referenced in the manuscript.

```{r example_alpha}
library(mpw)
## basic example code

Pf.time<- c(0,  14 , 28 , 42 , 56 , 70 , 84 , 98 , 112, 126, 140, 154, 168, 182, 196)
Pf.F1  <- c(0, .18,.19 ,.22, .25 , .27, .28, .34, .44, .50, .60, .72, .75, .81, .93)/100
Pf.F0  <- c(0, .29,.60, 1.00,1.38,1.75,2.25,2.97,3.50,4.25,4.94,5.53,6.00,6.31,6.94)/100

time <- Pf.time
F1 <- Pf.F1
F0 <- Pf.F0

tvec.in <- c(7, seq(14,max(time)-14,length.out=14),max(time)-7)
init.vals <- c(log(2.4), -10, rep( 0,length(tvec.in)))


## fit F_alpha for a given H_PARM
H_PARM <- 0.72
fit_F_alpha <- function(x){mean((popavg_dist(time, 
                                            knots= tvec.in, 
                                            logk0=x[1], 
                                            g0=x[2], 
                                            delta_vec=x[-c(1,2)],
                                            h_parm = H_PARM,
                                            frailty="PS")  - F0)^2) }


plac_fit_F_alpha <- 
  optim(init.vals, fit_F_alpha,
        method="Nelder-Mead",
        control=list(maxit=1e8))
print(plac_fit_F_alpha)

logk0p = plac_fit_F_alpha$par[1]
g0p = plac_fit_F_alpha$par[2]
delta_vec_p =plac_fit_F_alpha$par[-1*c(1,2)]
  
plot(time, 
     popavg_dist(time, knots=tvec.in, logk0=logk0p, g0=g0p, 
                 delta_vec = delta_vec_p, h_parm= H_PARM, frailty="PS"),
     type="l")
points(time, F0)

## plot the pop-avg (marginal) hazard:
plot(time,
     popavg_haz(time, knots=tvec.in, logk0=logk0p, g0=g0p,
                  delta_vec = delta_vec_p, h_parm= H_PARM,
                                            frailty="PS"),
     type="l", ylim=c(0,1e-3))

## plot the subject-specific hazard, fit in the presence
## of the h_parm value:

lines(time,
     subjspec_haz(time, knots=tvec.in, logk0=logk0p, g0=g0p,
                  delta_vec = delta_vec_p),
     lty=2)

```

## Gamma($\beta,\beta$) Example

This is the $F_{\beta}$ referenced in the manuscript.

```{r example_beta}
library(mpw)
## basic example code

Pf.time<- c(0,  14 , 28 , 42 , 56 , 70 , 84 , 98 , 112, 126, 140, 154, 168, 182, 196)
Pf.F1  <- c(0, .18,.19 ,.22, .25 , .27, .28, .34, .44, .50, .60, .72, .75, .81, .93)/100
Pf.F0  <- c(0, .29,.60, 1.00,1.38,1.75,2.25,2.97,3.50,4.25,4.94,5.53,6.00,6.31,6.94)/100

time <- Pf.time
F1 <- Pf.F1
F0 <- Pf.F0

tvec.in <- c(7, seq(14,max(time)-14,length.out=14),max(time)-7)
init.vals <- c(log(2.4), -10, rep( 0,length(tvec.in)))

## fit F_beta for a given H_PARM
H_PARM <- 0.72
fit_F_beta <- function(x){mean((popavg_dist(time, 
                                            knots= tvec.in, 
                                            logk0=x[1], 
                                            g0=x[2], 
                                            delta_vec=x[-c(1,2)],
                                            h_parm = H_PARM,
                                            frailty="GA")  - F0)^2) }


plac_fit_F_beta <- 
  optim(init.vals, fit_F_beta,
        method="Nelder-Mead",
        control=list(maxit=1e8))
print(plac_fit_F_beta)

logk0p = plac_fit_F_beta$par[1]
g0p = plac_fit_F_beta$par[2]
delta_vec_p =plac_fit_F_beta$par[-1*c(1,2)]
  
plot(time, 
     popavg_dist(time, knots=tvec.in, logk0=logk0p, g0=g0p, 
                 delta_vec = delta_vec_p, h_parm=H_PARM, frailty="GA"),
     type="l")
points(time, F0)

## plot the pop-avg (marginal) hazard:
plot(time,
     popavg_haz(time, knots=tvec.in, logk0=logk0p, g0=g0p,
                  delta_vec = delta_vec_p, h_parm= H_PARM, frailty="GA"),
     type="l", ylim=c(0,1e-3))

## plot the subject-specific hazard, fit in the presence
## of the h_parm value:

lines(time,
     subjspec_haz(time, knots=tvec.in, logk0=logk0p, g0=g0p,
                  delta_vec = delta_vec_p),
     lty=2)

```


## Inverse Gaussian($1,\lambda$) Example

This is the $F_{\lambda}$ referenced in the manuscript.

  * $F_{\lambda}(x) = 1-\exp[- \lambda ( (1+ \frac{2}{\lambda} \exp\eta )^{\frac{1}{2}} - 1) ]$
  
```{r example_lambda}
library(mpw)
## basic example code

Pf.time<- c(0,  14 , 28 , 42 , 56 , 70 , 84 , 98 , 112, 126, 140, 154, 168, 182, 196)
Pf.F1  <- c(0, .18,.19 ,.22, .25 , .27, .28, .34, .44, .50, .60, .72, .75, .81, .93)/100
Pf.F0  <- c(0, .29,.60, 1.00,1.38,1.75,2.25,2.97,3.50,4.25,4.94,5.53,6.00,6.31,6.94)/100

time <- Pf.time
F1 <- Pf.F1
F0 <- Pf.F0

tvec.in <- c(7, seq(14,max(time)-14,length.out=14),max(time)-7)
init.vals <- c(log(2.4), -10, rep( 0,length(tvec.in)))


## fit F_lambda for a given H_PARM
H_PARM <- 0.72
fit_F_lambda <- function(x){mean((popavg_dist(time, 
                                            knots= tvec.in, 
                                            logk0=x[1], 
                                            g0=x[2], 
                                            delta_vec=x[-c(1,2)],
                                            h_parm = H_PARM,
                                            frailty="IG")  - F0)^2) }


plac_fit_F_lambda <- 
  optim(init.vals, fit_F_lambda,
        method="Nelder-Mead",
        control=list(maxit=1e8))
print(plac_fit_F_lambda)

logk0p = plac_fit_F_lambda$par[1]
g0p = plac_fit_F_lambda$par[2]
delta_vec_p =plac_fit_F_lambda$par[-1*c(1,2)]
  
plot(time, 
     popavg_dist(time, knots=tvec.in, logk0=logk0p, g0=g0p, 
                 delta_vec = delta_vec_p, h_parm=H_PARM, frailty="IG"),
     type="l")
points(time, F0)
## plot the pop-avg (marginal) hazard:
plot(time,
     popavg_haz(time, knots=tvec.in, logk0=logk0p, g0=g0p,
                  delta_vec = delta_vec_p, h_parm=H_PARM, frailty="IG"),
     type="l", ylim=c(0,1e-3))

## plot the subject-specific hazard, fit in the presence
## of the h_parm value:

lines(time,
     subjspec_haz(time, knots=tvec.in, logk0=logk0p, g0=g0p,
                  delta_vec = delta_vec_p),
     lty=2)
```


## Two-Point($\rho$, $\xi$) Example

This is the $F_{\rho}$ referenced in the manuscript.

  * $F_{\rho}(x) =  1-             \left(           (1-\rho)  \exp[- \xi \exp\eta ] + \rho \exp[-\xi^{\prime} \exp \eta ] \right)$

```{r example_rho}
library(mpw)
## basic example code

Pf.time<- c(0,  14 , 28 , 42 , 56 , 70 , 84 , 98 , 112, 126, 140, 154, 168, 182, 196)
Pf.F1  <- c(0, .18,.19 ,.22, .25 , .27, .28, .34, .44, .50, .60, .72, .75, .81, .93)/100
Pf.F0  <- c(0, .29,.60, 1.00,1.38,1.75,2.25,2.97,3.50,4.25,4.94,5.53,6.00,6.31,6.94)/100

time <- Pf.time
F1 <- Pf.F1
F0 <- Pf.F0

tvec.in <- c(7, seq(14,max(time)-14,length.out=14),max(time)-7)
init.vals <- c(log(2.4), -10, rep( 0,length(tvec.in)))

## fit F_rho for a given H_PARM and XI <1
## (1-H_PARM) will have value XI < 1
##    H_PARM  will have XI_prime > 1 that ensures E[u] = 1
H_PARM <- 0.1
XI <- 1/10
fit_F_rho <- function(x){mean((popavg_dist(time, 
                                            knots= tvec.in, 
                                            logk0=x[1], 
                                            g0=x[2], 
                                            delta_vec=x[-c(1,2)],
                                            h_parm = H_PARM,
                                            frailty="TP1",
                                            xi = XI)  - F0)^2) }


plac_fit_F_rho <- 
  optim(init.vals, fit_F_rho,
        method="Nelder-Mead",
        control=list(maxit=1e8))
print(plac_fit_F_rho)

logk0p = plac_fit_F_rho$par[1]
g0p = plac_fit_F_rho$par[2]
delta_vec_p =plac_fit_F_rho$par[-1*c(1,2)]
  
plot(time, 
     popavg_dist(time, knots=tvec.in, logk0=logk0p, g0=g0p, 
                 delta_vec = delta_vec_p, h_parm=H_PARM, 
                 frailty="TP1", xi = XI),
     type="l")
points(time, F0)

## plot the pop-avg (marginal) hazard:
plot(time,
     popavg_haz(time, knots=tvec.in, logk0=logk0p, g0=g0p,
                  delta_vec = delta_vec_p, h_parm=H_PARM, 
                frailty="TP1", xi=XI),
     type="l", ylim=c(0,1e-2))

## plot the subject-specific hazard, fit in the presence
## of the h_parm value:

lines(time,
     subjspec_haz(time, knots=tvec.in, logk0=logk0p, g0=g0p,
                  delta_vec = delta_vec_p),
     lty=2)

```



## Two-Point($\omega$, $n$, $s$) Example

This is the $F_{\omega}$ referenced in the manuscript.

```{r example_omega}
library(mpw)
## basic example code

Pf.time<- c(0,  14 , 28 , 42 , 56 , 70 , 84 , 98 , 112, 126, 140, 154, 168, 182, 196)
Pf.F1  <- c(0, .18,.19 ,.22, .25 , .27, .28, .34, .44, .50, .60, .72, .75, .81, .93)/100
Pf.F0  <- c(0, .29,.60, 1.00,1.38,1.75,2.25,2.97,3.50,4.25,4.94,5.53,6.00,6.31,6.94)/100

time <- Pf.time
F1 <- Pf.F1
F0 <- Pf.F0

tvec.in <- c(7, seq(14,max(time)-14,length.out=14),max(time)-7)
init.vals <- c(log(2.4), -10, rep( 0,length(tvec.in)))


fit_F_omega <- function(x){mean((popavg_dist(time, 
                                            knots= tvec.in, 
                                            logk0=x[1], 
                                            g0=x[2], 
                                            delta_vec=x[-c(1,2)],
                                            h_parm = 0.50,
                                            frailty="TPU",
                                            n=0.25,
                                            s=4.00)  - F0)^2) }


plac_fit_F_omega <- 
  optim(init.vals, fit_F_omega,
        method="Nelder-Mead",
        control=list(maxit=1e8))
print(plac_fit_F_omega)

logk0p = plac_fit_F_omega$par[1]
g0p = plac_fit_F_omega$par[2]
delta_vec_p =plac_fit_F_omega$par[-1*c(1,2)]
  
plot(time, 
     popavg_dist(time, knots=tvec.in, logk0=logk0p, g0=g0p, 
                 delta_vec = delta_vec_p, h_parm=0.50, frailty="TPU",
                 n=0.25, s=4.00))
```

