---
title: "Weibull-Weibull-Positive Stable"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r data}
library(mpw)
## This data resembles that in Thomas et al
time<- c(0,14,28,42,56,70,84,98,112,126,140,154,168,182,196)
F1  <- c(0,.18,.19,.22,.25,.27,.28,.34,.44,.50,.60,.72,.75,.81,.93)/100
F0  <- c(0,.29,.60,1,1.38,1.75,2.25,2.97,3.50,4.25,4.94,5.53,6.00,6.31,6.94)/100

## plot the data points, 
## with interpolated lines
plot  (time, F0, type="l", col="black", xlab="Days", ylab="CDF")
points(time, F0, col="black", pch=15)
lines (time, F1, col="blue")
points(time, F1, col="blue", pch=16)
legend("topleft", c("Placebo", "Vaccine"),
       col=c("black","blue"),
       lty=1,
       pch=c(15,16)
)

```

Set the `h_parm` and `frailty` distribution for this example.

```{r setup}
H_PARM <- 0.72
FRAILTY <- "PS"
```
## Two knots

### Two step fitting procedure

#### 1. Fit for placebo group

#### 2. Fit for vaccine group

### Plot population avg CDF
  * Quick check of fit
  
### Plot population avg HR
  * uses parameters from two step fitting procedure
  * HR(0) = 1
  
### Add subject-specific HRs
  * should flatten

## Sixteen knots

```{r knot16_specific_setup}
tvec.in <- c(7, seq(14,max(time)-14,length.out=14),max(time)-7)
tvec.in
```

### Two step fitting procedure

#### 1. Fit for placebo group
```{r}
fit_F_alpha <- function(x){mean((popavg_dist(time, 
                                            knots= tvec.in, 
                                            logk0=x[1], 
                                            g0=x[2], 
                                            delta_vec=x[-c(1,2)],
                                            h_parm = H_PARM,
                                            frailty=FRAILTY)  - F0)^2) }

init.vals <- c(log(2.4), -10, rep( 0,length(tvec.in)))
plac_fit_F_alpha <- 
  optim(init.vals, fit_F_alpha,
        method="Nelder-Mead",
        control=list(maxit=1e8))
print(plac_fit_F_alpha)

logk0p = plac_fit_F_alpha$par[1]
g0p = plac_fit_F_alpha$par[2]
delta_vec_p =plac_fit_F_alpha$par[-1*c(1,2)]
```

#### 2. Fit for vaccine group

Now take `logk0p` and `g0p`, the parameters that control the first (leftmost)
piece for the placebo group and use them for `logk0v` and `g0v` when estimating the pieces for the vaccine group.  This forces the first pieces to be the same for both groups (i.e. *ramp-up time*).

```{r}
fit_F_alpha <- function(x){mean((popavg_dist(time, 
                                            knots= tvec.in, 
                                            logk0=logk0p, 
                                            g0=g0p, 
                                            delta_vec=x,
                                            h_parm = H_PARM,
                                            frailty=FRAILTY)  - F1)^2) }

init.vals <- rep( 0,length(tvec.in))
vacc_fit_F_alpha <- 
  optim(init.vals, fit_F_alpha,
        method="Nelder-Mead",
        control=list(maxit=1e8))
print(vacc_fit_F_alpha)

logk0v = logk0p
g0v = g0p
delta_vec_v =vacc_fit_F_alpha$par
```

### Plot population avg CDF
  * Quick check of fit

We use the knots and the original time in `time.dist` to do a quick check of the fit.  We put vertical dashed lines in the plot to denote knot placement.

```{r}
time.dist <- sort(c(time, tvec.in))
plac.dist <- popavg_dist(      x = time.dist, 
                           knots = tvec.in,
                           logk0 = logk0p,
                              g0 = g0p,
                       delta_vec = delta_vec_p,
                          h_parm = H_PARM,
                         frailty = FRAILTY)

vacc.dist <- popavg_dist(      x = time.dist, 
                           knots = tvec.in,
                           logk0 = logk0v,
                              g0 = g0v,
                       delta_vec = delta_vec_v,
                          h_parm = H_PARM,
                         frailty = FRAILTY)
  
plot  (time.dist, plac.dist, type="l", col="black", xlab="Days", ylab="CDF")
points(time     , F0       , col="black", pch=15)
lines (time.dist, vacc.dist, col="blue")
points(time     , F1, col="blue", pch=16)
abline(v=tvec.in, lty=2, col="grey")
legend("topleft", c("Placebo", "Vaccine"),
       col=c("black","blue"),
       lty=1,
       pch=c(15,16)
)

```

Note from Days `0` to `r tvec.in[1]` the curves are the same. 

### Plot population avg HR
  * uses parameters from two step fitting procedure
  * HR(0) = 1

We use the parameters of the previous fits to estimate the 
population average hazard, which we can fit with `mpw::popavg_haz()`.

```{r}
rnorm(1)

```
  
### Add subject-specific HRs
  * should flatten

