---
title: "Weibull-Weibull-Positive Stable"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true
---

# PS($\alpha,\alpha,0$) Positive Stable Frailty Example

This is the $F_{\alpha}$ referenced in the manuscript:

  * $F_{\alpha}(x) = 1-\exp[-\exp\alpha\eta]$

This is the PS($\alpha,\alpha,0$) Positive Stable Frailty density:

  * $f_{u_{i}}(u_{i} | \alpha) =  - \frac{1}{\pi u_{i}}  \sum_{k=1}^\infty \frac{\Gamma(k \alpha +1)}{k!}  (-u_{i}^{-\alpha})^k \sin{(\alpha k \pi)}$ 
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r data}
library(mpw)
## This data resembles that in Thomas et al
time<- c(0,14,28,42,56,70,84,98,112,126,140,154,168,182,196)
F1  <- c(0,.18,.19,.22,.25,.27,.28,.34,.44,.50,.60,.72,.75,.81,.93)/100
F0  <- c(0,.29,.60,1,1.38,1.75,2.25,2.97,3.50,4.25,4.94,5.53,6.00,6.31,6.94)/100

## plot the data points, 
## with interpolated lines
plot  (time, F0, type="l", col="black", xlab="Days", ylab="CDF")
points(time, F0, col="black", pch=15)
lines (time, F1, col="blue")
points(time, F1, col="blue", pch=16)
legend("topleft", c("Placebo", "Vaccine"),
       col=c("black","blue"),
       lty=1,
       pch=c(15,16)
)

```

Set the `h_parm` and `frailty` distribution for this example.

```{r setup}
H_PARM <- 0.72
FRAILTY <- "PS"
```

## Two knots
```{r knot02_specific_setup}
tvec.in <- c(5, 90+21)
tvec.in
```

### Two step fitting procedure

#### 1. Fit for placebo group
```{r}
fit_F_alpha <- function(x){mean((popavg_dist(time, 
                                            knots= tvec.in, 
                                            logk0=x[1], 
                                            g0=x[2], 
                                            delta_vec=x[-c(1,2)],
                                            h_parm = H_PARM,
                                            frailty=FRAILTY)  - F0)^2) }

init.vals <- c(log(2.4), -10, rep( 0,length(tvec.in)))
plac_fit_F_alpha <- 
  optim(init.vals, fit_F_alpha,
        method="Nelder-Mead",
        control=list(maxit=1e8))
print(plac_fit_F_alpha)

logk0p = plac_fit_F_alpha$par[1]
g0p = plac_fit_F_alpha$par[2]
delta_vec_p =plac_fit_F_alpha$par[-1*c(1,2)]
```

#### 2. Fit for vaccine group

Now take `logk0p` and `g0p`, the parameters that control the first (leftmost)
piece for the placebo group and use them for `logk0v` and `g0v` when estimating the pieces for the vaccine group.  This forces the first pieces to be the same for both groups (i.e. *ramp-up time*).

```{r}
fit_F_alpha <- function(x){mean((popavg_dist(time, 
                                            knots= tvec.in, 
                                            logk0=logk0p, 
                                            g0=g0p, 
                                            delta_vec=x,
                                            h_parm = H_PARM,
                                            frailty=FRAILTY)  - F1)^2) }

init.vals <- rep( 0,length(tvec.in))
vacc_fit_F_alpha <- 
  optim(init.vals, fit_F_alpha,
        method="Nelder-Mead",
        control=list(maxit=1e8))
print(vacc_fit_F_alpha)

logk0v = logk0p
g0v = g0p
delta_vec_v =vacc_fit_F_alpha$par
```

### Plot population avg CDF
  * Quick check of fit

We use the knots and the original time in `time.dist` to do a quick check of the fit.  We put vertical dashed lines in the plot to denote knot placement.

```{r}
time.dist <- seq(min(c(time, tvec.in)), max(c(time, tvec.in)), 0.1)
plac.dist <- popavg_dist(      x = time.dist, 
                           knots = tvec.in,
                           logk0 = logk0p,
                              g0 = g0p,
                       delta_vec = delta_vec_p,
                          h_parm = H_PARM,
                         frailty = FRAILTY)

vacc.dist <- popavg_dist(      x = time.dist, 
                           knots = tvec.in,
                           logk0 = logk0v,
                              g0 = g0v,
                       delta_vec = delta_vec_v,
                          h_parm = H_PARM,
                         frailty = FRAILTY)
  
plot  (time.dist, plac.dist, type="l", col="black", xlab="Days", ylab="CDF")
points(time     , F0       , col="black", pch=15)
lines (time.dist, vacc.dist, col="blue")
points(time     , F1, col="blue", pch=16)
abline(v=tvec.in, lty=2, col="grey")
legend("topleft", c("Placebo", "Vaccine"),
       col=c("black","blue"),
       lty=1,
       pch=c(15,16)
)
axis(1, at=c(tvec.in[1]))
```

Note from Days 0 to `r tvec.in[1]` the curves are the same. 

### Plot population avg HR
  * uses parameters from two step fitting procedure
  * HR(0) = 1

We use the parameters of the previous fits to estimate the 
population average hazard, which we can fit with `mpw::popavg_haz()`.
We adjust the time input so that the first element is just above 0.

```{r}
time.haz <- c(1e-4, time.dist[-1])
plac.haz <- popavg_haz(      x = time.haz, 
                           knots = tvec.in,
                           logk0 = logk0p,
                              g0 = g0p,
                       delta_vec = delta_vec_p,
                          h_parm = H_PARM,
                         frailty = FRAILTY)

vacc.haz <- popavg_haz(      x = time.haz, 
                           knots = tvec.in,
                           logk0 = logk0v,
                              g0 = g0v,
                       delta_vec = delta_vec_v,
                          h_parm = H_PARM,
                         frailty = FRAILTY)

plot  (time.haz, vacc.haz/plac.haz, type="l", col="purple", xlab="Days",
       ylab="Hazard Ratio", lwd=2)

abline(v=tvec.in, lty=2, col="grey")

axis(1, at=c(tvec.in[1]))

```
  
### Add subject-specific HRs
  * should flatten

The function `mpw::subjspec_haz()` does not require the arguments `frailty` or `h_parm` because it takes the inputs for the b,k, and delta pieces and models the subject-specific (aka conditional) hazard.  Doing this for the placebo group, then the vaccine group and dividing them gives the subject-specific hazard ratio (HR).

```{r}
plac.haz.ss <- subjspec_haz(      x = time.haz, 
                           knots = tvec.in,
                           logk0 = logk0p,
                              g0 = g0p,
                       delta_vec = delta_vec_p)

vacc.haz.ss <- popavg_haz(      x = time.haz, 
                           knots = tvec.in,
                           logk0 = logk0v,
                              g0 = g0v,
                       delta_vec = delta_vec_v)

plot  (time.haz, vacc.haz/plac.haz, type="l", col="purple", xlab="Days",
       ylab="Hazard Ratio", lwd=2)

axis(1, at=c(tvec.in[1]))

lines  (time.haz, vacc.haz.ss/plac.haz.ss, type="l", col="gold", xlab="Days",
        lwd=2, lty=2)
legend("topright", c("Population HR", "Subject-specific HR"),
       col=c("purple","gold"),
       lty=c(1,2),
       lwd=2
)


```

The plot above shows that for a positive-stable frailty with $\alpha$ = `r H_PARM` the subject-specific HR is flatter and lower than the population average HR.

## Sixteen knots

By increasing the knots, we get a "higher resolution" fit.  

```{r knot16_specific_setup}
tvec.in <- c(5, seq(14,max(time)-14,length.out=14),max(time)-7)
tvec.in
```

### Two step fitting procedure

#### 1. Fit for placebo group
```{r}
fit_F_alpha <- function(x){mean((popavg_dist(time, 
                                            knots= tvec.in, 
                                            logk0=x[1], 
                                            g0=x[2], 
                                            delta_vec=x[-c(1,2)],
                                            h_parm = H_PARM,
                                            frailty=FRAILTY)  - F0)^2) }

init.vals <- c(log(2.4), -10, rep( 0,length(tvec.in)))
plac_fit_F_alpha <- 
  optim(init.vals, fit_F_alpha,
        method="Nelder-Mead",
        control=list(maxit=1e8))
print(plac_fit_F_alpha)

logk0p = plac_fit_F_alpha$par[1]
g0p = plac_fit_F_alpha$par[2]
delta_vec_p =plac_fit_F_alpha$par[-1*c(1,2)]
```

#### 2. Fit for vaccine group

Now take `logk0p` and `g0p`, the parameters that control the first (leftmost)
piece for the placebo group and use them for `logk0v` and `g0v` when estimating the pieces for the vaccine group.  This forces the first pieces to be the same for both groups (i.e. *ramp-up time*).

```{r}
fit_F_alpha <- function(x){mean((popavg_dist(time, 
                                            knots= tvec.in, 
                                            logk0=logk0p, 
                                            g0=g0p, 
                                            delta_vec=x,
                                            h_parm = H_PARM,
                                            frailty=FRAILTY)  - F1)^2) }

init.vals <- rep( 0,length(tvec.in))
vacc_fit_F_alpha <- 
  optim(init.vals, fit_F_alpha,
        method="Nelder-Mead",
        control=list(maxit=1e8))
print(vacc_fit_F_alpha)

logk0v = logk0p
g0v = g0p
delta_vec_v =vacc_fit_F_alpha$par
```

### Plot population avg CDF
  * Quick check of fit

We use the knots and the original time in `time.dist` to do a quick check of the fit.  We put vertical dashed lines in the plot to denote knot placement.

```{r}
time.dist <- seq(min(c(time, tvec.in)), max(c(time, tvec.in)), 0.1)
plac.dist <- popavg_dist(      x = time.dist, 
                           knots = tvec.in,
                           logk0 = logk0p,
                              g0 = g0p,
                       delta_vec = delta_vec_p,
                          h_parm = H_PARM,
                         frailty = FRAILTY)

vacc.dist <- popavg_dist(      x = time.dist, 
                           knots = tvec.in,
                           logk0 = logk0v,
                              g0 = g0v,
                       delta_vec = delta_vec_v,
                          h_parm = H_PARM,
                         frailty = FRAILTY)
  
plot  (time.dist, plac.dist, type="l", col="black", xlab="Days", ylab="CDF")
points(time     , F0       , col="black", pch=15)
lines (time.dist, vacc.dist, col="blue")
points(time     , F1, col="blue", pch=16)
abline(v=tvec.in, lty=2, col="grey")
legend("topleft", c("Placebo", "Vaccine"),
       col=c("black","blue"),
       lty=1,
       pch=c(15,16)
)
axis(1, at=c(tvec.in[1]))
```

Note from Days 0 to `r tvec.in[1]` the curves are the same. 

### Plot population avg HR
  * uses parameters from two step fitting procedure
  * HR(0) = 1

We use the parameters of the previous fits to estimate the 
population average hazard, which we can fit with `mpw::popavg_haz()`.
We adjust the time input so that the first element is just above 0.

```{r}
time.haz <- c(1e-4, time.dist[-1])
plac.haz <- popavg_haz(      x = time.haz, 
                           knots = tvec.in,
                           logk0 = logk0p,
                              g0 = g0p,
                       delta_vec = delta_vec_p,
                          h_parm = H_PARM,
                         frailty = FRAILTY)

vacc.haz <- popavg_haz(      x = time.haz, 
                           knots = tvec.in,
                           logk0 = logk0v,
                              g0 = g0v,
                       delta_vec = delta_vec_v,
                          h_parm = H_PARM,
                         frailty = FRAILTY)

plot  (time.haz, vacc.haz/plac.haz, type="l", col="purple", xlab="Days", ylab="Hazard Ratio", lwd=2)

abline(v=tvec.in, lty=2, col="grey")

axis(1, at=c(tvec.in[1]))

```
  
### Add subject-specific HRs
  * should flatten

The function `mpw::subjspec_haz()` does not require the arguments `frailty` or `h_parm` because it takes the inputs for the b,k, and delta pieces and models the subject-specific (aka conditional) hazard.  Doing this for the placebo group, then the vaccine group and dividing them gives the subject-specific hazard ratio (HR).

```{r}
plac.haz.ss <- subjspec_haz(      x = time.haz, 
                           knots = tvec.in,
                           logk0 = logk0p,
                              g0 = g0p,
                       delta_vec = delta_vec_p)

vacc.haz.ss <- popavg_haz(      x = time.haz, 
                           knots = tvec.in,
                           logk0 = logk0v,
                              g0 = g0v,
                       delta_vec = delta_vec_v)

plot  (time.haz, vacc.haz/plac.haz, type="l", col="purple", xlab="Days",
       ylab="Hazard Ratio", lwd=2)

axis(1, at=c(tvec.in[1]))

lines  (time.haz, vacc.haz.ss/plac.haz.ss, type="l", col="gold", xlab="Days",
        lwd=2, lty=2)
legend("topright", c("Population HR", "Subject-specific HR"),
       col=c("purple","gold"),
       lty=c(1,2),
       lwd=2
)
```

The plot above shows that for a positive-stable frailty with $\alpha$ = `r H_PARM` the subject-specific HR is flatter and lower than the population average HR.

